Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-03-30 Write-ups/Manuscript.tex
data application use daily weigh-in data

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage[]{mathpazo}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
  pdfauthor={Ying Jin; Andrew Leroux},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 17, 2023}

@@

To address these problems, one may turn to functional mixed effect
models instead when measures are dense across the domain. Such methods
accommodate more flexible correlation structure by modeling
subject-specific random effects as a function, and non-parametric
smoothing (\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) can
be incorporated to speed up the computation, such as spline basis
functions or eigenfunctions from functional principal component analysis
(fPCA). The introduction of basis functions also makes out-of-sample
prediction more straightforward. Instead of estimating subject-specific
random effects of new observations, we can simply estimate
coefficients/loadings on the basis function used for smoothing. Existing
research on dynamic prediction with functional data analysis methods has
been focusing on continuous/Gaussian outcomes, modelling
subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors. However, little extension was
made on prediction of non-Gaussian functions, such as binary and count
outcomes.

@@

where g is a appropriate link function, \(\beta_0(t)\) is the population
mean of latent function, and \(b_i(t)\) is a subjects-level random
effect function follows a zero-mean Gaussian process.

While \(b_i(t)\) is not observed, it can be approximated under the FPCA
framework \(b_i(t)=\sum_{k=1}^K\xi_{ik}\phi_{ik}(t)+\epsilon_i(t)\).
Here \(\phi_{ik}(t)\) is a set of orthogonal eigenfunctions that
explains the most variation in \(b_i(t)\), and \(\xi_{ik}\) are
subject-specific PC scores (or loadings) on each eigenfunction.
Additionally, \(\xi_{ik}\) are mutually independent obver both subject
(\(i\)) and eigenfunctions (\(k\)). That is, each \(\xi_{ik}\) follows
normal distribution \(N(0, \lambda_k)\) where \(\lambda_k\) is the kth
eigenvalues: \(\int \phi_k^2(t)dt=\lambda_k\). \(\epsilon_i(t)\) here is
a residual function that accounts for the variation not explained by the
first K eigenfunctions from FPCA. We assume it follows a zero-mean
Gaussian process. At a specific point t,
\(\epsilon_i(t) \sim N(0, \sigma^2)\).

Based on the problem set up above, we propose the following algorithm
for PFCA on the unobserved latent process \(\eta_i(t)\):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t))] = \beta_0(s)+b_i(s)\) for all \(t\) in
  bin \(s\). From this series of models we can get estiamtes of
  population mean \(\hat{\beta}_0(s)\) and subject-level random effect
  \(\hat{b}_i(s)\), thus estimate of the individual latent functions at
  every bin: \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions \(\hat{\eta}_i(s)\), and
  obtain estimates of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), population mean
  \(\hat{\beta}_0(s)\) and residual variance \(\hat{\sigma}^2\).
\item
  With components extracted above, calculate the maximum likelihood
  estimate (MLE) of the subject-specific PC scores \(\hat{\xi}_{ik}\) of
  new samples based on their partially observed data. Then the value of
  latent functions at unobserved points can be estimated as
  \(\hat{\eta_i}(s)=\hat{\beta}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)\)
\end{enumerate}

Following the algorithm above, predictions of individual latent
functions are made on the binned grid based on partially observed
non-Gaussian functions tracks of new subjects. Since the bins are set up
to be small in length, the binned grid would still be dense enough.
However, in situations where we need predictions on the original,
un-binned grid instead, linear interpolation turns out to be a fast,
convenient way with good performance for prediction at points between
bins.

Precision of prediction, usually quantified by the variance of
prediction error \(Var(\hat{\eta_i}(s)-\eta_i(s))\), is also
straightforward under this framework. In step 4 we calculated the MLE of
\(\hat{\xi}_{ik}\). Based on likelihood theory, its variance can be
estimated with observed information \({I}(\hat{\xi}_{ik})\), which is
essentially the second derivative of likelihood at \(\hat{\xi}_{ik}\).
Therefore, the variation of prediction interval is:

\[Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2\]
Where \(\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))\) and
\(\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})\).

@@

In this section, we illustrate the predictive performance and
computational efficiency of the proposed method through a simulation
study. We simulated 50 datasets, each with 500 subjects. For every
subject, we generate 1000 binary outcomes \(Y_i(t) \in (0, 1)\) across
functional domain \(t \in [0, 1]\), where the distribution of outcome is
characterized by a continuous latent function. The data generation
mechanism can be expressed as follows:

\[\begin{aligned}
Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)
\end{aligned}\]

In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
independent normal random variables \(\xi_{ik}\sim N(0, \gamma_k)\).
Here we set the values of \(\gamma_k\) to be \(0.5^{k-1}\),
\(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
generate data on a regular grid, which means observations points are
equally distributed across \([0, 1]\) and are the same for all subjects.

We use two metrics to evaluate the out-of-sample predictive performance:
integrated squared error (ISE) and
Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
accucay of latent continuous function. It is evaluated on the binned
grid at midpoints of each unobserved bin. If the entire functional
domian has S bins, but we have observations up to the mth bin, then ISE
is defined as
\(\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
The second metric, AUC, focuses on evaluation of prediction of the
binary outcome. Since the binary outcomes are generated on the original,
un-binned grid, we evaluated AUC on this grid as well and estimated
values of latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1,
and Figure 1 presents the predicted function curves of four
randomly-drawn subjects from the first simulation. The prediction is
conditioning on different length of observed track (specifically, with
observations up to t = 200, 400, 600, 800 respectively), and evaluation
is made on equal-length time windows on the unobserved tracks following
the maximum observation time. As Table 1 reveals, the fGFPCA
outperformed GLMMadaptive under every scenario, which was expected since
GLMMadaptive can only accomodate simple model structures. In addition,
fGFPCA also takes less computation time. For one simulated dataset,
fGFPCA spent 4.94 minutes on model fitting and out-of-sample prediction,
while GLMMadaptive took 5.11 minutes.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/Figure-1.pdf}
\caption{Predicted track of four randomly selected subjects. The grey
dash line indicates the true latent continuous function. The dashed
lines indicated predicted latent function tracks, and color indicates
different observations time.}
\end{figure}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage[]{mathpazo}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
  pdfauthor={Ying Jin; Andrew Leroux},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 30, 2023}

@@

To address these problems, one may turn to functional mixed effect
models instead when measures are dense across the domain of measures.
Such methods accommodate more flexible correlation structure by modeling
subject-specific random effects as a function, and non-parametric
smoothing (\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) can
be incorporated to speed up the computation, such as spline basis
functions or eigenfunctions from functional principal component analysis
(FPCA). The introduction of basis functions also makes out-of-sample
prediction more straightforward. Instead of estimating subject-specific
random effects of new observations, we can simply estimate
coefficients/loadings on the basis function used for smoothing. Existing
research on dynamic prediction with functional data analysis methods has
been focusing on continuous/Gaussian outcomes, modelling
subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors.

@@

where g is an appropriate link function, \(\beta_0(t)\) is the
population mean of latent function, and \(b_i(t)\) is a subjects-level
random effect function following a zero-mean Gaussian process.

According to the Karhunen-Loève theorem, the unobserved random effect
function \(b_i(t)\) can be represented as the linear combination of a
infinite set of orthogonal functions:
\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). We approximate this
function by a finite set of \(\phi_{k}(t)\), such that
\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)\), where
\(\epsilon_i(t)\) indicates random noise introduced by this
approximation, following a zero-mean Gaussian process. At each specific
point t, \(\epsilon_i(t) \sim N(0, \sigma^2)\).

In practice, this approximation could be implemented through an FPCA
procedure on \(\eta_i(t)\). In this case
\(\phi_{k}(t),\ k \in \{1,...,K\}\) are the first K eigenfunctions that
explain the most variation in \(\eta_i(t)\), and \(\xi_{ik}\) are
subject-specific PC scores (or loadings) on each eigenfunction.
Additionally, \(\xi_{ik}\) are mutually independent over both subject
(\(i\)) and eigenfunctions (\(k\)). That is, each \(\xi_{ik}\) follows
normal distribution \(N(0, \lambda_k)\) where \(\lambda_k\) is the kth
eigenvalues.

\textbf{Fast Generalized FPCA} Based on the problem set up above, we
propose the following algorithm for fast, generalizable implementation
of FPCA (fGFPCA) on the unobserved latent process \(\eta_i(t)\):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t))] = \beta_0(s)+b_i(s)\) for all \(t\) in
  bin \(s\). From this series of models we can estimate the individual
  latent functions at every bin:
  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions:
  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
  and obtain estimates of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
  and residual variance \(\hat{\sigma}^2\).
\end{enumerate}

\textbf{Dynamic prediction} Now assume we have a new observations with
partially observed outcome \(Y_{new}(t)\), \(t\leq t_m\). The maximum
observed point is \(t_m\), which is a point along the functional domain
and belongs to bin \(s_m\) on the binned grid from step 1 above. With
components extracted from the FPCA model in step 3, we will be able to
make predictions on the future, unobserved bins. Specifically, with the
partially observed binary function \(Y_{new}(t)\), we will be able to
calculate the maximum likelihood estimate (MLE) of the subject-specific
PC scores \(\hat{\xi}_{ik}\). Then the value of latent functions at
unobserved points can be estimated as
\(\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)\)

Following the algorithm above, predictions of individual latent
functions are made on the binned grid based on partially observed
non-Gaussian functions tracks of new subjects. Since the bins are set up
to be small in length, the binned grid would still be dense. However, in
situations where predictions on the original, un-binned grid is needed,
linear interpolation turns out to be a fast, convenient way with good
performance for prediction at points between bins.

@@

In this section, we illustrate the predictive performance and
computational efficiency of the proposed method through a simulation
study. We simulated 50 datasets, each with 500 subjects. For every
subject, we generate 1000 binary outcomes \(Y_i(t) \in \{0, 1\}\) across
functional domain \(t \in [0, 1000]\), where the distribution of outcome
is characterized by a continuous latent function. The data generation
mechanism can be expressed as follows:

\[\begin{aligned}
Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
\end{aligned}\]

In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
independent normal random variables \(\xi_{ik}\sim N(0, \gamma_k)\).
Here we set the values of \(\gamma_k\) to be \(0.5^{k-1}\),
\(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
generate data on a regular grid, which means observations points are
equally distributed across \([0, 1000]\) and are the same for all
subjects.

We use two metrics to evaluate the out-of-sample predictive performance:
integrated squared error (ISE) and
Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
accuracy of latent continuous function. It is evaluated on the binned
grid at midpoints of each unobserved bin. If the entire functional
domain has S bins, but we have observations up to the mth bin, then ISE
is defined as
\(\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
The second metric, AUC, focuses on evaluation of prediction of the
binary outcome. Since the binary outcomes are generated on the original,
un-binned grid, we evaluated AUC on this grid and estimated values of
latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1,
and Figure 1 presents the predicted function curves of four
randomly-drawn subjects from the first simulation. The prediction is
conditioning on different length of observed track (specifically, with
observations up to t = 200, 400, 600, 800 respectively), and evaluation
is made on equal-length time windows on the unobserved tracks following
the maximum observation time. As Table 1 reveals, our proposed method
(fGFPCA) outperformed GLMMadaptive under every scenario, which was
expected since GLMMadaptive can only accommodate simple model
structures. In addition, fGFPCA also took less computation time. For one
simulated dataset, fGFPCA spent 4.94 minutes on model fitting and
out-of-sample prediction, while GLMMadaptive took 5.11 minutes.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/Figure-1.pdf}
\caption{Predicted track of four randomly selected subjects from one
simulated dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicated predicted latent
function tracks, and color indicates different observation time.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/Figure_appl-1.pdf}
\caption{Predicted track of four randomly selected subjects from the
daily weigh-in dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicated predicted latent
function tracks, and color indicates different observation time.}
\end{figure}

\caption{\label{tab:unnamed-chunk-3}AUC of fGFPCA and GLMM adaptive on the daily weigh-in dataset}
\centering
\begin{tabular}[t]{llccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Maximum observed time} \\
\cmidrule(l{3pt}r{3pt}){2-7}
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{fGFPCA} & \multicolumn{3}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
  & 100 & 200 & 300 & 100 & 200 & 300\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Prediction time window}}\\
\hspace{1em}(100, 200] & 0.75 &  &  & 0.76 &  & \\
\hspace{1em}(200, 300] & 0.77 & 0.82 &  & 0.76 & 0.80 & \\
\hspace{1em}(300, 400] & 0.76 & 0.81 & 0.84 & 0.74 & 0.78 & 0.82\\
\bottomrule
\end{tabular}
\end{table}
